Product Price Analysis Using Observational Data

### 2024-11-22 paper/paper.qmd
Modify to be more concise

---
title: "Product Price Analysis Using Observational Data"
subtitle: "Analysis of Price Trends, Minimal Correlation with Unit Pricing, and Potential Data Biases"
author: 
  - Yizhe Chen
  - Bo Tang
  - Qizhou Xie
thanks: "Code and data are available at: [https://github.com/YizheChenUT/Product_Prices.git](https://github.com/YizheChenUT/Product_Prices.git)."
date: today
date-format: long
abstract: "This report analyzes product pricing data to understand trends in current prices, changes over time, and the relationship between current price and unit price. The analysis reveals a very weak correlation between current price and unit price, indicating that other factors likely play a more significant role in determining product prices. This finding suggests that pricing strategies may be influenced by variables not captured in simple price-per-unit calculations. Understanding these patterns is essential for businesses and consumers alike, as it highlights the complexity of pricing decisions in the marketplace."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

Understanding how product prices are determined is essential for businesses, consumers, and policymakers alike. Pricing decisions impact consumer behavior, market competition, and ultimately the profitability of businesses. Despite the importance of pricing, the specific factors that influence product prices are often complex and may vary widely across industries and vendors. This report seeks to contribute to this understanding by analyzing observational price data from multiple vendors, specifically examining the relationship between current product price and unit price, as well as trends in price changes over time.

This study uses data obtained from Project Hammer by Jacob Filipp [@filipp], a large-scale dataset capturing product prices and other relevant attributes. The analysis is conducted using Structured Query Language (SQL) [@sql] to process and clean the data, while the statistical programming language R [@citeR] is employed to generate tables and graphs. The primary objective of this analysis is to identify price trends, calculate changes in price over time, and investigate whether there exists a meaningful correlation between current price and price per unit. A notable finding of this analysis is the extremely weak negative correlation between these two variables, suggesting that unit price has little to no direct impact on the overall product price.

The importance of this finding lies in its implications for understanding pricing behavior. The weak correlation indicates that other factors—such as brand reputation, demand fluctuations, and vendor-specific pricing strategies—may play a more significant role in determining product prices than unit cost alone. This insight is crucial for stakeholders who may otherwise rely heavily on unit price as an indicator of pricing strategy, highlighting the need for more nuanced approaches to price setting and market analysis.

The remainder of this report is structured as follows. @sec-data provides an overview of the data and details the measurement and calculation of key variables. @sec-results presents the main findings. @sec-discussion discusses key issues in the analysis, including correlation vs. causation, missing data, and sources of bias. Finally, @sec-conclusion concludes the report by summarizing the findings gained from the analysis, emphasizing the limitations due to data quality issues, and offering recommendations for future research to improve understanding of product pricing determinants.

@@

The dataset for this analysis originates from Project Hammer, curated by Jacob Filipp [@filipp], and contains key pricing information for various products across multiple vendors. The data includes critical variables such as current price, old price, and price per unit, allowing us to calculate basic statistics, examine price changes over time, and assess potential relationships between unit pricing and overall product price. This dataset offers a diverse view of pricing strategies by covering a range of products, vendors, and pricing behaviors, providing a valuable resource for understanding real-world pricing dynamics.

The data collected from various vendors captures a snapshot of product prices at specific points in time. Each entry represents an observation of a product’s pricing across these key variables, providing insights into both static and dynamic pricing practices. In terms of broader context, datasets that track longitudinal price changes across diverse vendors and product categories are rare, making this dataset a unique resource for exploring how pricing varies by vendor and product type. While alternative datasets could include retail pricing indices or specific vendor pricing records, they often lack the granularity or cross-vendor scope of Project Hammer, limiting their applicability for this type of multi-faceted analysis.

@@

- Old Price (old_price): The old price is the previous recorded price for each product. This variable enables us to calculate the percentage change in price over time, offering insights into price trends and volatility. For example, a reduction in old price compared to current price could indicate a promotional discount, while an increase might suggest adjustments due to supply constraints or increased demand. Tracking these historical prices provides a dynamic view of pricing strategies and changes over time.

@@

To ensure data integrity, the dataset was meticulously filtered and cleaned, particularly focusing on the current price and old price columns to remove any missing or invalid entries. This cleaning process involved handling null values and standardizing units where necessary to maintain consistency across entries. Such preprocessing steps allow for a more accurate and meaningful analysis by reducing noise and focusing on valid, comparable data points. Each entry in the dataset thus represents a refined record of a product’s price at two different time points, along with its unit price, translating real-world pricing phenomena into a structured format suitable for analysis.

@@

This analysis provides insights into product pricing trends and variability within the dataset, but it also underscores the importance of context when interpreting observational data. The correlation analysis showed a very weak relationship between current price and price per unit, suggesting that other factors are likely more influential in determining product prices. Additionally, missing data and potential sources of bias—such as selection, temporal, and measurement biases—highlight the limitations inherent in this dataset.

In conclusion, while the data offers valuable observations, these findings should be interpreted cautiously. Future analyses could benefit from a more comprehensive dataset that includes consistent reporting practices across vendors, longitudinal data to capture temporal trends, and efforts to minimize missing values. This would provide a more robust foundation for understanding the true determinants of product pricing.

>>>>

---
title: "Product Price Analysis Using Observational Data"
subtitle: "Analysis of Price Trends, Minimal Correlation with Unit Pricing, and Potential Data Biases"
author: 
  - Yizhe Chen
  - Bo Tang
  - Qizhou Xie
thanks: "Code and data are available at: [https://github.com/YizheChenUT/Product_Prices.git](https://github.com/YizheChenUT/Product_Prices.git)."
date: today
date-format: long
abstract: "This report analyzes product pricing data to examine trends, price changes, and the weak correlation between current price and unit price. The findings suggest that factors beyond unit price, such as demand or branding, significantly influence pricing strategies. Understanding these patterns highlights the complexity of marketplace pricing decisions for businesses and consumers."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

This study uses data obtained from Project Hammer by Jacob Filipp [@filipp], a large-scale dataset capturing product prices and other relevant attributes. The analysis is conducted using Structured Query Language (SQL) [@sql] to process and clean the data, while the statistical programming language R [@citeR] is employed to generate tables and graphs. The analysis reveals a very weak negative correlation between current price and unit price, suggesting that unit cost has little direct impact on product pricing.

The remainder of this report is structured as follows. @sec-data provides an overview of the data and details the measurement and calculation of key variables. @sec-results presents the main findings. @sec-discussion discusses key issues in the analysis, including correlation vs. causation, missing data, and sources of bias. Finally, @sec-conclusion concludes the report by summarizing the findings gained from the analysis, emphasizing the limitations due to data quality issues, and offering recommendations for future research.

@@

This analysis uses data from Project Hammer, curated by Jacob Filipp [@filipp], which includes key pricing variables such as current price, old price, and price per unit. These variables enable the calculation of basic statistics, price changes over time, and potential relationships between unit pricing and overall price. The dataset provides a snapshot of product prices from multiple vendors, capturing diverse pricing strategies and behaviors.

While alternative datasets, such as retail pricing indices, exist, they often lack the granularity and cross-vendor scope of Project Hammer. This dataset’s unique combination of breadth and detail makes it particularly valuable for exploring how pricing varies across products and vendors.

@@

- Old Price (old_price): The old price is the previous recorded price for each product. This variable enables us to calculate the percentage change in price over time, offering views into price trends and volatility. Tracking these historical prices provides a dynamic view of pricing strategies and changes over time.

@@

@@

This analysis highlights product pricing trends and variability while emphasizing the limitations of observational data. The weak correlation between current price and price per unit suggests that other factors, such as brand or demand, play a greater role in pricing. Missing data and biases further limit the dataset's reliability.

Future research should use more comprehensive datasets with consistent reporting, longitudinal tracking, and reduced missing values to better understand the determinants of product pricing.
